Parallel Multipoint Approximation Method for Large Scale Optimization Problems \thanks{This study was supported by the Russian Science Foundation, project No. 16-11-10150.}

### 2017-11-21 2018/PAVT_Toropov/PAVT2018_MAM.tex
Iteration number corrected
Translate to English

\documentclass[runningheads,a4paper]{llncs}

\usepackage{amssymb,amsmath}
\setcounter{tocdepth}{3}
\usepackage{graphicx}
\usepackage{enumitem}
\usepackage{url}

@@

\begin{abstract}
TBD

@@

\section{Introduction}
TBD

In problems with a large (in the order of hundreds) number of design variables, the multipoint approximation method (MAM \cite{Toropov1989,Toropov1992,ToropovFilatov1993}) proved to be efficient, e.g. in turbomachinery applications \cite{ShahparPolynkinToropov2008,PolynkinToropovShahpar2008,PolynkinToropovShahpar2010} . This method is an iterative optimization technique based on mid-range approximations built in trust regions. A trust region is a sub-domain of the design space in which a set of design points, produced according to a small-scale design of experiments (DoE), are evaluated. These and a subset of previously evaluated design points are used to build metamodels of the objective and constraint functions that are considered to be valid within a current trust region. The trust region will then translate and change size as optimization progresses. The trust region strategy has gone through several stages of development to account for the presence of numerical noise in the response function values \cite{KeulenToropovMarkine1996,ToropovKeulenMarkine1996} and occasional simulation failures \cite{ToropovMarkineHolden1999}. The mid-range approximations used in the trust regions, as originally suggested in \cite{Toropov1989} for structural optimization problems, are intrinsically linear functions (i.e. nonlinear functions that can be led to a linear form by a simple transformation) for individual sub-structures, and assembly of them for the whole structure. This was enhanced by the use of gradient-assisted metamodels \cite{ToropovFilatov1993}, use of simplified numerical models that is also termed a multi-fidelity approach,\cite{ToropovMarkine1996} and the use of analytical models derived by Genetic Programming \cite{ToropovAlvarez1998}. One of the recent developments \cite{PolynkinToropov2012} involved the use of approximation assemblies, i.e. a two stage approximation building process that is conceptually similar to the original one used in \cite{Toropov1989} but is free from the limitation that lower level approximations are linked to individual substructures.

@@

TBD

@@

It is useful to start with a brief description of MAM. A typical formulation of a constrained optimization problem that MAM works with is as follows:
\begin{equation}
  \label{eq:problem}
  \begin{array}{c}
  \min\limits_{a_i \le x_i \le b_i}F_0(x) \\
  s.t.\; F_j(x) \le 1,\; j=1,\dots ,M,
  \end{array}
\end{equation}
where $x$ is the vector of design variables, $a$ and $b$ are the lower and upper bounds for the design variables, respectively, $F_0(x)$ is the objective function, and $F_j(x)$ are the constraints. The numbers of design variables and constraints are $n$ and $M$, respectively. MAM attempts to solve this problem by using approximations of the objective function and constraints in a series of trust regions. The trust region strategy seeks to zoom in on the region where the constrained minimum is achieved. It aims at finding a trust region that is sufficiently small for the approximations to be of sufficiently good quality to improve the design, and that contains the point of the constrained minimum as its interior point. The main loop of the MAM is organized as follows.

Algorithm (MAM).
\begin{enumerate}
\item Initialization: choose a starting point $ x^0$ and initial trust region $[ a^0,  b^0]$ such that $ x^0 \in [ a^0,  b^0]$.
\item At the $k$-th iteration the current approximation to the constrained minimum is $ x^k$, the current trust region is $[ a^k,  b^k] \subset [ a^0,  b^0]$.
  \begin{enumerate}[label=(\alph*)]
    \item Design of Experiments (DoE): a set of points $ x_k^i \in [ a^k,  b^k]$ is chosen to be used for approximation building. Responses are evaluated at the DoE points and approximations are built using the obtained values. Currently, the pool of approximation methods available in MAM consists of a metamodel assemblies \cite{PolynkinToropov2012} and the moving least-squares metamodels \cite{LancasterSalkauskas1981,Liszka1984,ChoiYounYang2001,ToropovSchrammSahaiJones2005}. Other metamodel types could be used as well.

@@

\section{Parallel Algorithm}
\label{sec:par_alg}

@@

@@

>>>>

\documentclass[runningheads,a4paper]{llncs}
\usepackage{amssymb,amsmath}
\setcounter{tocdepth}{3}
\usepackage{graphicx}
\usepackage{enumitem}
\usepackage{url}

\usepackage[T2A]{fontenc}
\usepackage[utf8x]{inputenc}
\usepackage[russian,english]{babel}

@@

\begin{abstract}
The paper presents a new development in the Multipoint Approximation Method (MAM) that makes it capable of handling large scale problems. The approach relies on approximations built in the space of design variables within the iterative trust-region-based framework of MAM. 
With the purpose of solving the problems of high dimensionality in a reasonable time, a parallel variant of the Multipoint Approximation Method (PMAM) has been developed. It was supposed that the values of the objective function and of the constraints were computed using the distributed memory (on several cluster nodes), whereas the optimization module run on a single node using the shared memory. The numerical experiments have been carried out on a benchmark example of structural optimization.

@@

\section{Introduction}

\Russian
In the present paper, the multipoint approximation method (MAM \cite{Toropov1989,Toropov1992,ToropovFilatov1993}) and its application to the large scale optimization problems are considered. In problems with a large (in the order of hundreds) number of design variables MAM proved to be efficient, e.g. in turbomachinery applications \cite{ShahparPolynkinToropov2008,PolynkinToropovShahpar2008,PolynkinToropovShahpar2010}. This method is an iterative optimization technique based on mid-range approximations built in trust regions. A trust region is a sub-domain of the design space in which a set of design points, produced according to a small-scale design of experiments (DoE), are evaluated. These and a subset of previously evaluated design points are used to build metamodels of the objective and constraint functions that are considered to be valid within a current trust region. The trust region will then translate and change size as optimization progresses. The trust region strategy has gone through several stages of development to account for the presence of numerical noise in the response function values \cite{KeulenToropovMarkine1996,ToropovKeulenMarkine1996} and occasional simulation failures \cite{ToropovMarkineHolden1999}. The mid-range approximations used in the trust regions, as originally suggested in \cite{Toropov1989} for structural optimization problems, are intrinsically linear functions (i.e. nonlinear functions that can be led to a linear form by a simple transformation) for individual sub-structures, and assembly of them for the whole structure. This was enhanced by the use of gradient-assisted metamodels \cite{ToropovFilatov1993}, use of simplified numerical models that is also termed a multi-fidelity approach,\cite{ToropovMarkine1996} and the use of analytical models derived by Genetic Programming \cite{ToropovAlvarez1998}. One of the recent developments \cite{PolynkinToropov2012} involved the use of approximation assemblies, i.e. a two stage approximation building process that is conceptually similar to the original one used in \cite{Toropov1989} but is free from the limitation that lower level approximations are linked to individual substructures.

@@

This paper presents a Parallel Multipoint Approximation Method that makes it capable of handling problems with the number of design variables in the order of thousands. The parallel variant of the algorithm (with the use of the shared memory) has been developed with the purpose of minimizing the worktime of the part of the algorithm related to constructing the approximation and solving the approximated problem and not related to computing the values of the objective function and  constraints. The processes of computing the values of the objective function and constraints (with the use of the 
distributed memory) are supposed to be parallelized already.

@@

It is useful to start with a brief description of MAM. A typical formulation of a constrained optimization problem that MAM works with is as follows:
\begin{equation}
  \label{eq:problem}
  \begin{array}{c}
  \min\limits_{a_i \le x_i \le b_i}F_0(x) \\
  s.t.\; F_j(x) \le 1,\; j=1,\dots ,M,
  \end{array}
\end{equation}
where $x$ is the vector of design variables, $a$ and $b$ are the lower and upper bounds for the design variables, respectively, $F_0(x)$ is the objective function, and $F_j(x)$ are the constraints. The numbers of design variables and constraints are $n$ and $M$, respectively. MAM attempts to solve this problem by using approximations of the objective function and constraints in a series of trust regions. The trust region strategy seeks to zoom in on the region where the constrained minimum is achieved. It aims at finding a trust region that is sufficiently small for the approximations to be of sufficiently good quality to improve the design, and that contains the point of the constrained minimum as its interior point. The main loop of the MAM is organized as follows.
\\

\textbf{Algorithm (MAM).}
\begin{enumerate}
\item Initialization: choose a starting point $ x^0$ and initial trust region $[ a^0,  b^0]$ such that $ x^0 \in [ a^0,  b^0]$.
\item At the $k$-th iteration the current approximation to the constrained minimum is $ x^k$, the current trust region is $[ a^k,  b^k] \subset [ a^0,  b^0]$.
  \begin{enumerate}[label=(\alph*)]
    \item Design of Experiments (DoE): a set of points $ x_k^i \in [ a^k,  b^k]$ is chosen to be used for approximation building. Responses are evaluated at the DoE points and approximations are built using the obtained values. Currently, the pool of approximation methods available in MAM consists of a metamodel assemblies \cite{PolynkinToropov2012} and the moving least-squares metamodels \cite{LancasterSalkauskas1981,Liszka1984,ChoiYounYang2001,ToropovSchrammSahaiJones2005}. Other metamodel types could be used as well.

@@

\section{Parallel Multipoint Approximation Method}\label{sec:par_alg}

Let us consider possible methods of parallelization, which could be applied for the considered problems.

First, one can parallelize the computing of the functions describing the optimized object. This way is obvious as well as necessary one since in the industrial design optimization problems the computing of even a single function value might take several hours. However, this method is a specific one for each particular problem. Here the issues of computations are addressed at the level of the application software, in which the industrial modeling is performed (e.g. Ansys, OpenFOAM, etc.).

Second, one can correct the algorithm with the purpose of parallel computing several values of the objective function and constraints at different points of the search domain. According to the MAM rules, in design of experiments $P$ 
sampling points are formed in current trust region at every iteration. The function values at these points can be computed on different processors (nodes) in parallel. The above corresponds to the parallelization of Step 2a of the algorithm with the use of the distributed memory. The number of sampling points generated within each iteration may be set equal to $P = k\cdot NP$ number, where $NP$ is a number of available processors (or nodes) and $k \geq 1$. In  terms of the time, the latter will be equivalent to $NP$ function evaluations per step. This method has been implemented successfully in \cite{PolynkinToropov2008} and has demonstrated a good efficiency since at the number of the design variables of the order of $10^2$ major time is consumed by the function evaluations, the work of the MAM itself (in the sequential regime) introduces minor overhead.

However, when the number of variables becomes of the order of $10^3$, the work of the sequential part of MAM begins to affect the total problem solving time essentially (assuming the time of computing the objective function 
values remains constant). Thus, for the problem considered in Section 4 the time of execution of a single iteration of the method increased more than 2000 times (from 4 up to 9000 seconds) when increasing the number of variables from 100 up to  1000. Therefore, another approach to the parallelization of the algorithm has been applied within the framework of the 
present study: namely, the computational rules of MAM  providing the construction of the approximating, the solving of the approximated problem, and the choice of the next trust region (Steps 2b, 2c, and 2d of the algorithm) have been parallelized.

In order to find the most time-consuming parts of the sequential program developed earlier, we have applied Intel VTune Amplifier XE. The performed analysis has shown the matrix multiplication and solving SLAE when constructing the approximations as well as solving the approximating problem by SQP method to be the most time-consuming operations in the execution of MAM. The matrix multiplication and solving SLAE are standard operations implemented in many high-performance libraries. Here we used the corresponding parallel methods from Intel MKL library. We have parallelized the SQP method ourselves using OpenMP.

@@

Next, let us compare the worktime of the sequential and parallel algorithms when solving the large scale problems. 
The dimensionality of the problem being solved $N$ was varied from 100 up to 1000 design variables that corresponds to 50 to 500 segments of the cantilevered beam. The number of points in a trust region used to build the approximations was $1.5 N$. 
Both algorithms was run on a single node of the cluster (the specifications of the node are listed below), the parallel algorithm employed all 16 processors cores available. Since the time of computing of the objective 
function and constraints in the test problem being solved was negligible, these ones were computed on the same node.

Table 1 reflects the worktime (in seconds) of the sequential algorithm and of the parallel one subject to the number of variables. In Table 2, the number of MAM iterations $I_{MAM}$ as well as the number of the function evaluations $F_{MAM}$ required for solving the problem are presented. For comparison, the number of the function value computations, which would be required to solve the initial (non-approximated) problem by SQP method $F_{SQP}$ is given also. In all conducted experiments, the objective function values in the sequential and parallel versions of the algorithm were the same (up to the accuracy of the computational error) and differed from the solution obtained by SQP negligibly. The reduction of the number of the function evaluations required to solve the problem using MAM as compared to the use of SQP was demonstrated 
visibly.

\begin{table}
	\caption{Time and speedup}
	\label{tab:1}
	\center
	\begin{tabular}{cccc}
		\hline\noalign{\smallskip}
		$N$ & MAM time & PMAM time & Speedup \\
		\hline\noalign{\smallskip}
		100 & 48 &  34 &   \\
		200 & 48 &  34 &   \\
		500 & 48 &  34 &   \\
		1000 & 107479  &  19469 &   \\
		\noalign{\smallskip}\hline
	\end{tabular}
\end{table}

\begin{table}
	\caption{Function evaluations}
	\label{tab:2}
	\center
	\begin{tabular}{cccc}
		\hline\noalign{\smallskip}
		$N$ & $I_{MAM}$  & $F_{MAM}$ & $F_{SQP}$ \\
		\hline\noalign{\smallskip}
		100 & 10 &  2201 & 9901 \\
		200 & 10 &  4402 & 28143 \\
		500 &  &   &  \\
		1000 &  &   &  \\
		\noalign{\smallskip}\hline
	\end{tabular}
\end{table}

Computational experiments were carried out on a high-performance cluster of Lobachevsky State University of Nizhny Novgorod. The cluster node included two Intel Sandy Bridge E5-2660 2.2 GHz CPUs and 64 Gb RAM. The CPU had 8 cores, i.e. a total of 16 physical cores were available in the node. MS Visual Studio 15 and Intel Fortran Compiler were used to implement the algorithm.

@@

Recent developments in the Multipoint Approximation Method (MAM) made it capable of solving large scale industrial optimization problems. The fact that MAM solves the initial problem by using the approximations of the objective function and constraints is the primary distinctive feature of the method. Within the framework of the present study, the parallel version of MAM oriented onto the reduction of the worktime of the optimization algorithm (assuming that computing the values of the objective function and constraints has been parallelized already) has been developed. The performed experiments have demonstrated an acceptable speedup when solving the large scale problems employing 16 cores on a single cluster node. The performance was demonstrated on a benchmark example of structural 
optimization known as the scalable cantilevered beam.
